Personalised Map Matched Imputation: an approach to impute missing data in smartphone location logs

### 2018-04-29 manuscript/manuscript.Rmd
APA manuscript

---
title: "Manuscript"
author: "Boaz Sobrado"
output:
  pdf_document: default
  html_document: default
---

@@

Smartphone location measures are obtained primarily (but not exclusively) by Global Positioning System (GPS) measurements. A GPS sensor uses the distance between a device and several satellites to determine the location of the device. Although using a GPS sensors is the most accurate way to establish location on a smartphone, the GPS sensor is the most energy consuming sensor on most smartphones [@lamarca_place_2005; @chen_practical_2006]. In order to avoid battery depletion and to overcome computational constraints smartphones also use less-accurate heuristics such as WiFi access points and cellphone tower triangulation. Smartphone location logs contain measurements from all of these sources, usually in the form of timestamped lattitude, longitude and accuracy values. The accuracy $a$ of any given measurement is given in meters such that it represents the radius of a 67% confidence circle [@location_history_timeline_2017]. In other words, the true location of a device should be within the radius $a$ of the measurement 67% of the time.

@@

include_graphics("img/missingBoaz5minExample.png")
#change y axis lables to natural numbers, x-axis lables to shorter periods and colour to shapes.

@@

```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement of a morning journey on February 15th 2017. This includes all measurements from midnight to midday. The red circles denote the accuracy of all logged measurement points (the raw data). The points connected in time are connected by a line. The blue line shows the path without the most inaccurate (accuracy > 400 meters) points filtered out. The red line shows the path with all measurements included. @palmius_detecting_2017 note that in smartphone custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour. Inaccurate measures are often followed by more accurate measures. There are several recurring low-accuracy points, such as the one in the northwest corner, possibly the result of cell-phone tower triangulation.",out.width = '100%'}
include_graphics("img/journeyTillMiddayBoaz.png")
```

@@

GPS measurements provide us with coordinates on the surface of the earth, which is ellipsoid shaped. Projecting three dimensional measurements in $\mathbb{R}^3$ onto a two dimensional plane in $\mathbb{R}^2$ results in distortion. For clarity, when we use the term distance we refer to the geodesic distances on an ellipsoid using the WGS84 ellipsoid parameters.

@@

1. Map building: Extract from measurements $D$ all pause locations and path clusters to create a personalised map.
2. Binning: Assign each measurement $D$ to a unique discrete location. 
3. Imputing: Use a classification method to predict missing measurements based on the time as well as the distance in time to the next and previous measurements.

@@

Following @wu's spatial-temporal-semantic (STS) feature extraction algorithm our aim is to transform pause and path locations into machine learning friendly discrete location sequences. There are multiple ways of extracting such measurement clusters in the literature, such as Spatio-Temporal Density-Based Spatial Clustering of Applications with Noise (ST-DBSCAN) and sequence oriented clustering (SOC) [@xiang;]. We will focus on two methods which explicitly with mobility patterns from unevenly sampled smartphone logs [ @palmius_detecting_2017 ;@barnett_inferring_2016]. Both of these methods pre-process the data and subsequently use two steps to extract pause locations: first they extract pauses and their corresponding locations, then they cluster pause locations based on spatial proximity. For simplicity this implementation of PMMI uses a stricter version of @barnett_inferring_2016's approach to extract pauses. However, there is no reason why more sophisticated methods cannot be used.

@@

This set of points were then hierarchically clustered using a distance matrix, such that all points within $d$ meters of each other were clustered into a pause location. Each pause location is a cluster.

For all remaining points we assume that they belong to paths. In this implementation of PMMI we use the following algorithm to estimate paths:

@@

At this point all empirically observed path clusters and pause clusters are extracted. However, there may be some overlap between pause clusters and path clusters. Thus, the clusters are clustered again, such that the pause clusters retain priority. This means that if a pause cluster and a path cluster are within less than $d$ meters of each other, the path cluster was removed. The reasoning for this is that the threshold for not being in a pause cluster should be higher, as individuals spend the majority of time at a pause cluster. The end result is a discretised map which contains pause and flight clusters based on the history of the individual.

### Binning

@@

The key modification in PMMI is that whilst @wu uses data from outside the persons location logs, we use the total location history of the individual to create a personalised map which will be then used to bin measurements. This is feasible for two reasons: humans tend to have repetitive movement habits and secondary logs tend to be long. To put it in simpler terms, we consider each measurement at $D_x$ as a sample of $Path_{ab}$, and by aggregating many measurements we can use them to map out $Path_ab$.

Thus, all measurements $D$ were are then assigned to a discrete bin on the personal map. This includes previous measurements which were discarded from the map building excercise due to an accuracy $a$ value which exceeded $a_{\text{P lim}}$ or $a_{\text{F lim}}$. Again, for simplicity we used a simple assigning function, whereby the measurements where assigned to the bin nearest to the measurement.

@@

1. The previous and subsequent observed bin.
2. The distance in time to the next & previous bin.
3. The time of the day encoded as a cyclical two-dimensional feature.
4. The day of the month as a binary class matrix.
5. The month of the year as a binary class matrix.

For the encoding of the time of day we took the cosine and the sine transforms of the amount of seconds that have elapsed after midnight [@london]. This is necessary so that the model can understand that one second past midnight and one second before midnight are in fact two seconds away from each other.

@@

For simplicity, we only used a time period when the indivdual was living in the Netherlands in the model. Analyses as well as implementations and adaptions @palmius_detecting_2017's and @barnett_inferring_2016's model were performed using R and a multitude of other statistical packages [@base;@ggplot2;@dplyr;@sp1;@sp2]. **add packages** All the code is available on a public repository.

@@

1. Evaluating the performance of the map building and assigning functions.
2. Comparing the performance of PMMI using a) baseline models b) performance with randomly removed data in comparison to the aforementioned methods [@palmius_detecting_2017; @barnett_inferring_2016] and c) objective ground-truth data.

@@

In order to evalute the map building and binning we will first visually evaluate the paths and pause locations. A visual evaluation of paths superimposed on is an established way to heuristically check their accuracy (*add reference*). Then, let the average distance between the actual measured point and the binned point be the *deviation distance* $\delta_{\text{dev}}$. With respect to the deviation distance $\delta_{\text{dev}}$, we expect:

@@

We will compare the peformance of PMMI, @palmius_detecting_2017 and  @barnett_inferring_2016. In addition, we will also compute a naive model, which simply imputes as the missing value the previous observed value. To compare the performance of these methods we will remove 25% of measured time intervals at random within a four week period. We will make our comparisons for intervals of 5 minutes, 1 hour and 1 day. In other words, we will remove 25% of time intervals at random, while varying the size of the time intervals removed.

@@

In addition, other measures we will report for PMMI are accuracy (in what percentage of the cases was the appropriate cluster predicted) and the expected distance. The expected distance is the crossproduct of the estimated probabilities that the individual is at a given cluster and the distances between the clusters to the true cluster. For example, if the true location of an individual is cluster A (cluster A is 10 meters away from cluster B) and the model assigns a probability of 0.9 at cluster A and 0.1 at cluster B, then the expected distance would be 1 meter.

Finally, we will take objective real world data and compare it to predicted values. We will use information from the Dutch public transportation card as objective real world data. The Dutch public transportation card provides users with timestamped data of when and where they board, change lines or leave public transportation.

@@

The selection of these parameters was more-or-less heuristic.

When selecting parameters there is a trade off is between resolution and accuracy. This is because an increase in resolution comes at the expense of accuracy. For instance, by increasing the clustering distance parameter $d_{\text{Pause lim}}$ we can extract more valid pause locations at the expense of falsely categorising certain measurements to the wrong cluster. This is illustrated in Figures \@ref(fig:pauseLocUithof1) and \@ref(fig:pauseLocUithof2).

```{r, pauseLocUithof1,echo = FALSE,fig.cap="Example of pause locations in De Uithof using 150 meters and 400 meters as clustering parameters respectively.",out.width = '100%'}
#fix width
include_graphics("img/mapPauseRes1.png")
```

```{r, pauseLocUithof2,echo = FALSE,fig.cap="Example of pause locations in De Uithof using 150 meters and 400 meters as clustering parameters respectively.",out.width = '100%'}
include_graphics("img/mapPauseRes2.png")
```

The end result of step 3 of MYMETHOD results in a personalised map with pause and path clusters. An excerpt can be seen in Figure \@ref(fig:mapClusterMap). As expected, MYMETHODs path extraction yields greater accuracy for frequently ocurring paths. It is important to remind the reader that despite the close overlap with features on the map,  such as roads and railway lines, MYMETHOD is map agnostic and uses no information from the map to recreate these features.

```{r, mapClusterMap,echo = FALSE,fig.cap="Excerpt of the cluster map if an individual. Red points are pause locations, blue points are path locations.",out.width = '100%'}
#fix width
include_graphics("img/mapClusterMap.png")
```

For the entire period when the individual is in the Netherlands, there is a mean deviance of 40 meters and a median deviance of 15 meters. Around 69% of the deviance values are within their corresponding accuracy value, which is close to the theoretical $\frac{2}{3}$ value that is expected.

For the narrower period of March 2017, figure \@ref(fig:devianceComp) illustrates the relationship between the deviance and accuracy. Approximately 74% of the deviance values are within their corresponding accuracy values, which is close to the theoretical $\frac{2}{3}$ value that is expected. The mean and median deviance are 38 and 14 meters respectively.

```{r, devianceComp,echo = FALSE,fig.cap="Excerpt. MISSING CAPTION HERE",out.width = '100%'}
#fix width
# add figure with deviance comparisons three columns each method once
```

In comparison, Palmius' method has a median deviance of 3 meters, with a mean of 115 due to high deviance outliers. Over 90% of deviance values are within accuracy, which suggests that the model is binning noise. In Figure \@ref(fig:palmiusVme) the key difference is visible. The Palmius method is much more sensitive to inaccurate values, such as the one at the top right of the figure.

On the other hand, Barnett's method has mean deviance of 343 meters and a median deviance of 8 meters. Barnett's deviance is neccessarily higher than Palmius', as they downsample and subsequently aggregate into pauses and linear flights.

```{r, palmiusVsMe1,echo = FALSE,fig.cap="Excerpt. MISSING CAPTION HERE",out.width = '100%'}
#fix width
include_graphics("img/palmiusVme.png")
```

```{r, palmiusVsMe2,echo = FALSE,fig.cap="Excerpt. MISSING CAPTION HERE",out.width = '100%'}
include_graphics("img/binComp.png")
```

### Comparison to alternative models

```{r,resulttable, echo = F}
library(knitr)
library(kableExtra)

resultsDf <- data.frame( mean5min = c(82,43,269,426),
                         median5min = c(4,0,0,0),
                         meanexpDist5min = c(NA,NA,414,NA),
                         medianexpDist5min = c(NA,NA,0,NA),
                         mean1hr = c(345,497,908,1502),
                         median1hr = c(6,4,0,0),
                         meanexpDist1hour = c(NA,NA,916, NA),
                         medianexpDis1hour = c(NA,NA,0, NA),
                         mean1day = c(9273,NA,5757,14266),
                         median1day = c(12,NA,0,1288),
                         meanexpDist1Day = c(NA,NA,5757, NA),
                         medianexpDis1Day = c(NA,NA,0, NA))

rownames(resultsDf) <- c("Barnett & Onella", "Palmius", "Sobrado","Naive")
colnames(resultsDf) <- c("Mean","Median","Expected Mean","Expected Median",
                         "Mean","Median","Expected Mean","Expected Median",
                         "Mean","Median","Expected Mean","Expected Median")

# median 12 meters mean 9273
library(knitr)
library(kableExtra)
options(knitr.kable.NA = '-')
kable(resultsDf, "html") %>%
  kable_styling("striped") %>%
  add_header_above(c(" " = 1, "Five Minutes" = 4, "One Hour" = 4,"One Day" = 4))

@@

In terms of coverage, it is important to note that the Palmius model failed to impute 3% of all removed values for both the 5 minute and over 10% for the 1 hour tests. Palmius' model failed to impute a single value for the day tests. Barnett's failed to impute 11% of the missing values. MYMETHOD made an imputation for all missing values.

As for MYMETHOD's accuracy in this period, the prediction accuracy was 88%, 73% and 47% for the 5 minute, 1 hour and 1 day periods respectively. As a comparison, the naive model's accuracy ratings were 87%, 68% and 24% respectively.

@@

The naive model reaches an accuracy of 16%, with a median distance of 555 meters and mean distance of 3303 meters. On the other hand MYMETHOD has an accuracy of 24%, with a median distance of 1037 meters and a median distance of 5637. The expected distance has a mean of 6432 and a median of 1037.

@@

### Effect on aggregate measures

Social scientists are most interested in aggregating spatiotemporal data to more socially relevant information, such as distance travelled. As we discussed earlier, aggregations without data processing can be highly biased. However, as an example we calculate time spent at home of the user for each day of the week in the month of February (\@ref(fig:aggrePlot)). Between individuals, time spent at home has been found to be a reliable predictor of extraversion [@harari_using_2016].

```{r, aggrePlot,echo = FALSE,fig.cap="Proportion of time spent at home in February 2017. We estimate this by calculating the mean lattitude and longitude for every 5 minute time period in the month. Missing values are filled in by the previous observed value. Behavioural trends are evident, the user spends more time at home on weekends than on weekdays. Moreover, the variance is greater on weekends, due to travel.",out.width = '100%'}
include_graphics("img/aggPlot.png")
```

## Discussion

Overall the MYMETHOD performed well, although not substantially better than the naive model.  In addition, the comparison to the performance of the Barnett&Onella and Palmius models is somewhat unfair, as they were created for custom logs, not secondary logs. Nonetheless, the comparison remains valid as they are the closest we found to a missing data imputation.

In addition to higher accuracy under the conditions of secondary logs, the advantages of MYMETHOD are increased coverage and flexibility for missing data imputation, robustness to irregular sampling, the ability to model complex non-linear interactions in its imputations, and the ability to use historical records to smoothen movement noise.

MYMETHOD's increased coverage and flexibility comes from its ability to make complex non-linear predictions. For instance, in a given missing period it might make sense to predict that the individual is either at home, or at the office, or at a shop with equal probability. While MYMETHOD can make such an imputation, none of the alternative methods can do this. Moreover, the ability to take the prediction probability values from the neural network also helps in dealing with uncertainty. A known-drawback of single imputation is that it takes an imputed value and treats it as observed (*RUBIN*). Simple rule-based methods such as Palmius' are essentially algorithmic single imputation methods. With MYMETHOD it is possible to construct confidence intervals using the predicted probabilities of each estimate.

With respect to irregular sampling, alternative methods use temporally based downsampling in order to reduce noise. This leads to deterioration in resolution not only over space, but also over time. A combination of irregular sampling with the fluctuating accuracy values can lead to nonsensical results. For instance, consider a case where there are two inaccurate measurements in movement at 12:00:01 and 12:04:59. Downsampling over 5 minute periods will lead to a value that will be the mean of the two inaccurate samples, which is likely to be a location the individual is certainly not. MYMETHOD instead downsamples spatially, which ensures that the binned location is one which is composed of the mean of hundreds of observations.

While both @barnett_inferring_2016 and @palmius_detecting_2017 use historical data to smoothen pause locations by clustering pause locations with with a close degree of spatial proximity, neither of them do the same for paths. Hence all paths are calculated based on sensor input only. This may be feasible with high frequency regularly sampled short duration logs but creates noise with secondary logs. Moreover, with secondary logs it is feasible to spatially "average out" multiple samples of the same path in order to recreate it in its entirety. For instance, although the mean sampling frequency during train travels on the Amsterdam-Utrecht line is low (*report value here*) the personalised map manages to recreate the train line almost perfectly, despite being completely map agnostic.

@@

Furthermore, MYMETHOD also has several limitations. The pause extracting function, the path clustering function, the classification model and the assignment function remain simplistic and could be improved. Indeed, our intention is not to develop an optimal classification model or any of the other components of MYMETHOD, but rather to show a new way in which secondary logs can be made useful by social scientists.

With regards to pause extraction, the probability of a pause at a given location is certainly related to other factors, such as the time of the day as well as the prior history of pauses at that location. These factors are not taken into account in the pause extraction function. Improved methods would do well to do so.

As for the path clustering function, a drawback of the current method is that the density of the clusters is a function of the clustering parameter $d$, the distance between the observed points and their sampling density. It does not take into account the length of the path as well as the average sampling frequency of the path. This is an issue because it can lead to clusters to which data points are seldomly assigned. For example, while the Amsterdam-Utrecht line has been mapped out almost perfectly, many of the clusters along the route have only been assigned few measurements. This leads to difficulties in the machine learning part of the model, as unfrequently observed clusters are hard for the model to predict.

Similarly, a limitation of the current assignment function is that it only takes into account the distance between the observed measurement and all the clusters when assigning each measurement to a cluster. It does not take into account any contextual information that can be gleaned from the entire movement history of the information, such as what path they are on. For instance, assume that it is known that an individual is travelling from point A to point B along path AB, and there is an inaccurate measurement closest to a cluster which belongs to path AC. By only taking distance into account, the measurement can get assigned to the wrong cluster on path AC. An improvement would be use a Bayesian method, whereby assignment is a function of both the measurement and the ANN model. This could take the form of a Kalman filter. In this case the state side of the model would represent a probabilistic representation of where the individual could be based on what the ANN has extracted from the individuals movement history. The space side of the model would be a measurement equation representing the measurement and the uncertainty surrounding it in the form of $a$.

As for the simplicity of the neural network which was used to generate predictions. A more sophisticated recurrant neural network (RNN), or a long short-term memory recurrant neural network (LSTM) would likely perform significantly better.

## Conclusion

Advantages of MYMETHOD:

>>>>

---
title: "Personalised Map Matched Imputation: an approach to impute missing data in smartphone location logs"
author: "Boaz Sobrado"
output:
  pdf_document: default
  html_document: default
---

@@

Smartphone location measures are obtained primarily (but not exclusively) by Global Positioning System (GPS) measurements. A GPS sensor uses the distance between a device and several satellites to determine the location of the device. Although using a GPS sensor is the most accurate way to establish location on a smartphone, the GPS sensor is also the most energy consuming sensor on most smartphones [@lamarca_place_2005; @chen_practical_2006]. In order to avoid battery depletion and to overcome computational constraints smartphones also use less-accurate heuristics such as WiFi access points and cellphone tower triangulation. Smartphone location logs contain measurements from all of these sources, usually in the form of timestamped lattitude, longitude and accuracy values. The accuracy $a$ of any given measurement is given in meters such that it represents the radius of a 67% confidence circle [@location_history_timeline_2017]. In other words, the true location of a device should be within the radius $a$ of the measurement 67% of the time.

@@

include_graphics("img/missin5minExample2018.png")
```

@@

```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement of a morning journey on February 15th 2017. This includes all measurements from midnight to midday. The red circles denote the accuracy of all logged measurement points (the raw data). The points connected in time are connected by a line. The blue line shows the path without the most inaccurate (accuracy > 400 meters) points filtered out. The red line shows the path with all measurements included. P note that in smartphone custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour. Inaccurate measures are often followed by more accurate measures. There are several recurring low-accuracy points, such as the one in the northwest corner, possibly the result of cellphone tower triangulation.",out.width = '100%'}
include_graphics("img/journeyTillMiddayBoaz.png")
```

@@

Location measurements, such as those produced by GPS sensors, provide us with coordinates (lattitude and longitude) on the surface of the earth, which is ellipsoid shaped. Projecting three dimensional measurements in $\mathbb{R}^3$ onto a two dimensional plane in $\mathbb{R}^2$ results in distortion. For clarity, when we use the term distance we refer to the geodesic distances on an ellipsoid using the WGS84 ellipsoid parameters.

@@

1. *Map building*: Extract from measurements $D$ all pause location bins and path locatin bins to create a personalised map.
2. *Binning*: Assign each measurement $D$ to a unique discrete location bin. 
3. *Imputing*: Use a classification method to predict missing measurements based on all the availble information.

@@

Following @wu's spatial-temporal-semantic (STS) feature extraction algorithm our aim is to transform pause and path locations into machine learning friendly discrete location sequences. There are multiple ways of extracting such measurement clusters in the literature, such as Spatio-Temporal Density-Based Spatial Clustering of Applications with Noise (ST-DBSCAN) and sequence oriented clustering (SOC) [@xiang;]. We will focus on two methods which explicitly with mobility patterns from unevenly sampled smartphone logs [ @palmius_detecting_2017 ;@barnett_inferring_2016]. Both of these methods pre-process the data and subsequently use two steps to extract pause locations: first they extract pauses and their corresponding locations, then they cluster pause locations based on spatial proximity. This implementation of PMMI uses a stricter version of @barnett_inferring_2016's approach to extract pauses.

@@

This set of points were then hierarchically clustered using a distance matrix, such that all points within $d$ meters of each other were clustered into a pause location. Each pause location is a bin.

For all remaining measurements we assume that they belong to paths. In this implementation of PMMI we use the following algorithm to estimate paths:

@@

At this point all empirically observed path bins and pause bins are extracted. However, there may be some overlap between pause bins and path bins. Thus, the bins are clustered again, such that the pause bins retain priority. This means that if a pause bin and a path bin are within less than $d$ meters of each other, the path bin is removed. The reasoning for this is that the threshold for not being in a pause cluster should be higher, as individuals spend the majority of time at a pause cluster. The end result is a discretised map which contains pause and flight bins based on the entire log history of the individual.

#### Binning

@@

PMMI uses a similar logic, but without using any external geographic database. The key modification in PMMI is that whilst @wu uses a map from outside the persons location logs, we use the total location history of the individual to create a personalised map. This map is subsequently used to bin measurements. This is feasible for two reasons: humans tend to have repetitive movement habits and secondary logs tend to be long. To put it in simpler terms, we consider each measurement at $D_x$ as a sample of $Path_{ab}$, and by aggregating many measurements we can use them to map out $Path_{ab}$.

Thus, all measurements $D$ were are then assigned to a discrete bin on the personal map. This includes previous measurements which were discarded from the map building excercise due to an accuracy $a$ value which exceeded $a_{\text{P lim}}$ or $a_{\text{F lim}}$. In this implementation we used a simple assigning function, whereby the measurements where assigned to the bin nearest to the measurement.

@@

1. The previous and subsequent observed bin as a binary class matrix.
2. The distance in time to the next & previous bin.
3. The time of the day encoded as a cyclical two-dimensional feature.
4. The day of the month as a binary class matrix.
5. The month of the year as a binary class matrix.

For the encoding of the time of day we took the cosine and the sine transforms of the amount of seconds that have elapsed after midnight [@london]. This is necessary so that the model can understand that one second past midnight and one second before midnight are in fact two seconds away from each other.   Moreover we scaled the non-binary values to occupy a range between 0 and 1 in order to ensure convergence.

@@

For simplicity, we only used a time period when the indivdual was living in the Netherlands in the model. This leaves us with 156000 measurements over a period of less than six months. Analyses as well as implementations and adaptions @palmius_detecting_2017's and @barnett_inferring_2016's model were performed using R and a multitude of other statistical packages [@base;@ggplot2;@dplyr;@sp1;@sp2]. All the code is available on a public repository [@mygithub].

@@

1. Evaluating the performance of the map building and assigning functions.
2. Comparing the performance of PMMI using a) baseline models b) performance with randomly removed data in comparison to the aforementioned methods [@palmius_detecting_2017; @barnett_inferring_2016] and c) objective ground-truth data (public transportation timestamps).

@@

In order to evalute the map building and binning we will first visually evaluate the paths and pause locations. A visual evaluation of paths superimposed on is an established way to heuristically check their accuracy (e.g. @lcm). Then, let the average distance between the actual measured point and the binned point be the *deviation distance* $\delta_{\text{dev}}$. With respect to the deviation distance $\delta_{\text{dev}}$, we expect:

@@

We will compare the peformance of PMMI, @palmius_detecting_2017 and @barnett_inferring_2016. In addition, we will also compute a naive model, which simply imputes as the missing value the previous observed value. The naive model will serve as a baseline model. To compare the performance of these methods we will remove 25% of measured time intervals at random within a four week period. We will make our comparisons for intervals of 5 minutes, 1 hour and 1 day. In other words, we will remove 25% of time intervals at random, while varying the size of the time intervals removed.

For the @barnett_inferring_2016 and PMMI models we will use all the available data to train the models with the exception of the time periods being investigated. @palmius_detecting_2017's model does not require training.

@@

In addition, other measures of interest for PMMI are accuracy (in what percentage of the cases was the appropriate cluster predicted), the *confidence* and the *distance expectation*. The confidence is the probability with which the model predicts the most likely cluster. For instance, if the model predicts the missing bin to be bin A with a probability of 0.9 then we can say it has high confidence in the prediction. Similarly, the distance expectation is the crossproduct of the estimated probabilities that the individual is at any of all given clusters and the distances between the clusters to the true cluster. For example, if the true location of an individual is bin A (bin A is 10 meters away from bin B) and the model assigns a probability of 0.9 at bin A and 0.1 at bin B, then the distance expectation would be 1 meter.

Finally, we will take objective real world data and compare it to predicted values. We will use information from the Dutch public transportation card. The Dutch public transportation service provides users with timestamped data of when and where they board, change lines or leave public transportation. To be able to make a comparison between models we will remove all measurements from within the 5 minute period that a timestamped measurement is available. Then we will use each model to impute the location of the individual within that period.

@@

The selection of these parameters was more-or-less heuristically driven based on their on their ability to extract a meaningful personalised map. When selecting parameters there is a trade off is between bias and precision. This is because an increase in precision in the form of a higher for high density locations resolution comes at the expense of precision as assigning measurements to bins becomes more difficult. For instance, by increasing the clustering distance parameter $d_{\text{Pause lim}}$ we can extract more valid pause locations at the expense of falsely categorising certain measurements to the wrong cluster. This is illustrated in Figure \@ref(fig:pauseLocUithof1).

```{r, pauseLocUithof1,echo = FALSE,fig.cap="Example of pause locations in De Uithof university campus using 150 meters (blue) and 400 meters (red) as clustering parameters.",out.width = '100%'}
#fix width
include_graphics("img/distanceParam.png")
```

Map building results in a personalised map with pause and path clusters. An excerpt can be seen in Figure \@ref(fig:mapClusterMap). It is important to remind the reader that PMMI is map agnostic and uses no information from the map. Therefore, the close overlap with features on the map, such as pause bins at relevant buildings and transportation clusters, as well as the flight bins following roads and railway lines indicate a high degree of precision in personal map building. As expected, PMMI's path extraction yields greater accuracy for frequently ocurring paths. For example, the frequently travelled Amsterdam-Utrecht railway line has been extracted almost perfectly, while the less frequently travelled Utrecht-Enschede line is far sparser.

```{r, mapClusterMap,echo = FALSE,fig.cap="Excerpt of the cluster map of an individual. Red points are pause locations, blue points are path locations.",out.width = '100%'}
#fix width
include_graphics("img/mapClusterMap.png")
```

For the entire period examined period the we find a deviance of 40 meters and a median deviance of 15 meters. Around 69% of the deviance values are within their corresponding accuracy value, which is close to the theoretical $\frac{2}{3}$ value that is expected. Approximately 9% of values were not taken into account when creating the bins,given that their the accuracy $a$ exceeded $a_{\text{Path Lim}}$.

For the narrower period of March 2017 approximately 74% of the deviance values are within their corresponding accuracy values, which is close to the theoretical $\frac{2}{3}$ value that is expected. The raw unweighted mean and median deviance are 38 and 14 meters respectively. If we weight them such that each 5 minute interval is weighted equally to ease comparison with the other two methods, we find a mean deviance of 35 meters with a median of 12.8.

In comparison, Palmius' method has a median deviance $\delta_{\text{dev}}$ of 3 meters, with a mean of 115 due to high deviance outliers. On the other hand, Barnett's method has mean deviance $\delta_{\text{dev}}$ of 343 meters and a median deviance of 8 meters. Barnett's deviance is neccessarily higher than Palmius', as they downsample temporally (like Palmius) and subsequently aggregate into pauses and linear flights.

The key difference between temporal and spatial downsampling is shown in Figure \@ref(fig:palmiusVme1). Temporal downsampling is much more sensitive to noise in sparsely measured periods because it averages out values within five minute periods. Often there are only a few noisy measurements in those periods (see Figure \@ref(fig:longMeasurementsPerDay) ), which leads to a noise in the downsampled values. Unsurprisingly, there is a positive relationship between deviance and the amount of measurements in each downsampled interval. The fact that over 90% of deviance values are within accuracy (much higher than the expected theoretical 67%) confirms that temporal downsampling is not sufficiently filtering out the noise.

```{r, palmiusVsMe1,echo = FALSE,fig.cap="Excerpt. MISSING CAPTION HERE",out.width = '100%'}
#fix width
include_graphics("img/tempVsspace.png")
```

### Imputation evaluation

The @palmius_detecting_2017 model failed to impute 3% of all removed values for both the 5 minute and over 10% for the 1 hour tests. @palmius_detecting_2017' model failed to impute a single value for the day tests.
Similarly, @barnett_inferring_2016's method failed to impute over 11% of the missing values. PPMI made an imputation for all missing values.

```{r,resulttable, echo = F}
# results df

resultsDf <- data.frame( mean5min = c(82,43,269,426),
                         median5min = c(4,0,0,0),
                         mean1hr = c(345,497,908,1502),
                         median1hr = c(6,4,0,0),
                         mean1day = c(9273,NA,5757,14266),
                         median1day = c(12,NA,0,1288)
                         )

rownames(resultsDf) <- c("Barnett & Onella", "Palmius", "Sobrado","Naive Baseline")
colnames(resultsDf) <- c("Mean","Median",
                         "Mean","Median",
                         "Mean","Median")

# median 7 meters mean 60 not imputed 282 NA's Ian all data

library(knitr)
library(kableExtra)
options(knitr.kable.NA = '-')
kable(resultsDf, "latex") %>%
  kable_styling(latex_options = "striped") %>%
  add_header_above(c(" " = 1, "Five Minutes" = 2, "One Hour" = 2,"One Day" = 2))

@@

As for PPMI's accuracy in this period, the prediction accuracy was 88%, 73% and 47% for the 5 minute, 1 hour and 1 day periods respectively. As a comparison, the naive model's accuracy ratings were 87%, 68% and 24% respectively. The distance expectation values were very similar to the distance scores, albeit approximatley 5% higher. Confidence scores ranged from 0.006 to 1.

@@

Once we removed the measurements within the 5 minute period of each timestamp, we employed the models to predict the location of the individual at the timestamped period. The naive model reaches an accuracy of 16%, with a median distance of 555 meters and mean distance of 3303 meters. On the other hand MYMETHOD has an accuracy of 24%, with a median distance of 1037 meters and a mean distance of 5637. Again the expectation of the distance was quite similar to the distance with a mean of 6432 and a median of 1037 meters.

@@

### Example: effect on aggregate measures

Social scientists are most interested in aggregating spatiotemporal data to more socially relevant information, such as the amount of time spent at home. As an example we calculated the time spent at home of the user in the month of March (\@ref(fig:aggrePlot)) without any model and with all three of the investigated methods.

```{r, aggrePlot,echo = FALSE,fig.cap="Proportion of time spent at home in March 2017. The raw values are estimated by downsampling temporally the lattitude and longitude for every 5 minute time period in the month. We used each method's own binning method and classified as at home if the downsampled measurement was within 250 meters from home.",out.width = '100%'}

include_graphics("img/timeUse.png")
```

Interestingly, all three models suggest that the user spent approximately 60% of their time at home. However, without using any form of missing data imputation approximately 12% of the time is unaccounted for. Given that the amount of time spent at home has been found to be a reliable predictor of extraversion [@harari_using_2016] and the onset of depressive episodes in bipolar patients [@palmius_detecting_2017] obtaining an accurate value for mobility metrics like this is highly important to social scientists.

## Discussion & Conclusion

Overall the PMMI performed better than the alternative models, particularly during longer missing periods and with objective data. However, it did not perform substantially better than the naive baseline model. In addition, the comparison to the performance of the Barnett&Onella and Palmius models is somewhat unfair, as they were created for custom logs, not secondary logs. Nonetheless, the comparison remains valid as they are the closest we found to a missing data imputation methods in smartphone GPS logs.

In addition to higher accuracy under the conditions typical of secondary logs, the advantages of PPMI are increased coverage and flexibility for missing data imputation, robustness to irregular sampling, the ability to model complex non-linear interactions in its imputations, and the ability to use historical records to smooth movement noise.

PPMI's increased coverage and flexibility comes from its ability to make complex non-linear predictions. For instance, in a given missing period it might make sense to predict that the individual is either at home, or at the office, or at a shop with equal probability. While PPMI can make such an imputation, none of the alternative methods can do this. Moreover, the ability to take the prediction probability values from the neural network also helps in dealing with uncertainty. A known-drawback of single imputation is that it takes an imputed value and treats it as observed. Simple rule-based methods such as Palmius' are essentially algorithmic single imputation methods. With PPMI it is possible to model uncertainty using the predicted probabilities of each estimate. For instance, in the previous example, we could choose to only take estimates with a high degree of confidence, thus creating confidence intervals by adding and substracting the amount of cases where the location of the individual is ambiguous.

With respect to irregular sampling, alternative methods use temporally based downsampling in order to reduce noise. This leads to deterioration in resolution not only over space, but also over time. A combination of irregular sampling with the fluctuating accuracy values can lead to nonsensical results. For instance, consider a case where there are two inaccurate measurements in movement at 12:00:01 and 12:04:59. Downsampling over 5 minute periods will lead to a value that will be the mean of the two inaccurate samples, which is likely to be a location the individual is certainly not. PPMI instead downsamples spatially, which ensures that the binned location is one which is composed of the mean of hundreds of observations, not just the few that happen within a single period.

While both @barnett_inferring_2016 and @palmius_detecting_2017 use historical data to smoothen pause locations by clustering pause locations with with a close degree of spatial proximity, neither of them do the same for non-pause locations. This may be feasible with high frequency, regularly sampled short duration logs but creates noise with secondary logs. Moreover, with secondary logs it is feasible to spatially "average out" multiple samples of the same path in order to recreate it in its entirety. For instance, although the mean sampling frequency during train travels on the Amsterdam-Utrecht line is low (about 0.01 Hz) the personalised map manages to recreate the train line almost perfectly, despite being completely map agnostic.

@@

Furthermore, PPMI can be further developed. The map building function, the assignment function and the classification model remain simplistic and could be improved.

In map building, the probability of a pause at a given location is certainly related to other factors, such as the time of the day as well as the prior history of pauses at that location. These factors are not taken into account in the pause extraction function. Improved methods would do well to do so. As for paths, a drawback of the current method is that the density of the clusters is a function of the clustering parameter $d$, the distance between the observed points and their sampling density. It does not take into account the length of the path as well as the average sampling frequency of the path. This is an issue because it can lead to bins to which data points are seldomly assigned. For example, while the Amsterdam-Utrecht line has been mapped out almost perfectly, many of the clusters along the route have only been assigned few measurements. This leads to difficulties in the classification part of the model, as unfrequently observed clusters are hard for the model to predict.

The current assignment function simply assigns each masurement to the nearest bin. It does not take into account any contextual information that can be gleaned from the entire movement history of the information, such as what path they are on. For instance, assume that it is known that an individual is travelling from point A to point B along path AB, and there is an inaccurate measurement closest to a cluster which belongs to path AC. By only taking distance into account, the measurement can get assigned to the wrong cluster on path AC. An improvement would be use a Bayesian method, whereby assignment is a function of both the measurement and a model of the individuals movement history. In terms of state-space models like the Kalman filter the state equation would represent a probabilistic representation of where the individual could be based on the individuals entire movement history. The space side of the model would be a measurement equation representing the measurement and the uncertainty surrounding it in the form of $a$.

As for the simplicity of the classification method, the neural network which was used to generate predictions used no information on sequence patterns longer than the previous and next bin. A more sophisticated recurrant neural network (RNN), or a long short-term memory recurrant neural network (LSTM) would likely perform significantly better.
